---
title: "Multiple Regression"
author: "Dr. Hannay"
date: "November 28, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Multiple Regression
Looking at the colors in the above plot corresponding to different neighborhoods we can see that we might get a better model if we built a linear model for each neighborhood individually. If we include more than one explanatory variable this is called *multiple regression*. In general it is a bad idea to include too many explanatory variables, but in this case we are justified in including neighborhood information as the plot shows the neighborhoods behaving very differently. 

To allow for nieghborhood effects in our model we simply add this value to the model. 
```{r}
res.house.neigh<-lm(SalePrice.log10~Square.Feet.log10+Neighborhood, data=house)
summary(res.house.neigh)
```

The results here show that some neighborhoods have inherently more expensive houses. The coefficients for each neighborhood are added to the y-intercept value of 3.08. For example, the houses in the StoneBr neighborhood are best modeled by $$ \log_{10}(Price)=3.08+0.31+0.63 \log_{10} (sqft)=3.39+ 0.63 \log_{10}(sqft).$$


```{r}
house.StoneBr<-subset(house, house$Neighborhood=='StoneBr')
res.stonebr<-lm(SalePrice.log10~Square.Feet.log10, data=house.StoneBr)
summary(res.stonebr)
```
